---
title: "Homework 2"
author: "Sarah Younes"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

As always, I will begin by loading the tidyverse, and I will additionally load `readxl` for Problem 2.

```{r tidyverse, message = FALSE}
library(tidyverse)
library(readxl)
```

# Problem 1

First, I will clean the data in pols-month.csv.

```{r pols_month_df}
pols_month_df =
  read.csv("./Data/fivethirtyeight_data/pols-month.csv") |>
  separate(mon, into = c("year", "month", "day"), sep = "-") |>
  mutate(
     month = replace(month, month == "01", "Jan"),
     month = replace(month, month == "02", "Feb"),
     month = replace(month, month == "03", "Mar"),
     month = replace(month, month == "04", "Apr"),
     month = replace(month, month == "05", "May"),
     month = replace(month, month == "06", "Jun"),
     month = replace(month, month == "07", "Jul"),
     month = replace(month, month == "08", "Aug"),
     month = replace(month, month == "09", "Sep"),
     month = replace(month, month == "10", "Oct"),
     month = replace(month, month == "11", "Nov"),
     month = replace(month, month == "12", "Dec")
  ) |>
  mutate(
    president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")) |>
  mutate(
    year = as.integer(year)
  ) |>
  select(year, month, everything(), -day, -starts_with("prez"))
```
    
Second, I will clean the data in snp.csv.

```{r snp_df}
snp_df =
  read.csv("./Data/fivethirtyeight_data/snp.csv") |>
  separate(date, into = c("month", "day", "year"), sep = "/") |>
  mutate(
     month = replace(month, month == "1", "Jan"),
     month = replace(month, month == "2", "Feb"),
     month = replace(month, month == "3", "Mar"),
     month = replace(month, month == "4", "Apr"),
     month = replace(month, month == "5", "May"),
     month = replace(month, month == "6", "Jun"),
     month = replace(month, month == "7", "Jul"),
     month = replace(month, month == "8", "Aug"),
     month = replace(month, month == "9", "Sep"),
     month = replace(month, month == "10", "Oct"),
     month = replace(month, month == "11", "Nov"),
     month = replace(month, month == "12", "Dec")
  ) |>
  mutate(
    year = as.integer(year)
  ) |>
  select(year, month, close)
```
   
Third, I will tidy the data in unemployment.csv.

```{r unemployment_df}
unemployment_df =
  read.csv("./Data/fivethirtyeight_data/unemployment.csv") |>
  rename(year = Year) |>
  pivot_longer(
    Jan:Dec,
    names_to = "month",
    values_to = "unemployment"
  ) |>
  mutate(month = str_to_title(month))
```

Now, I will join the three datasets into a single data frame.

```{r merge problem 1}
problem_1_df =
  left_join(pols_month_df, snp_df) |>
  left_join(x = _, y = unemployment_df)
```

# Problem 2

First, I will load the `readxl` package and import and clean the Mr. Trash Wheel dataset. Data cleaning for this dataset involves setting missing values, cleaning the variable names, renaming two existing variables, creating two new variables, and changing the column type for one existing variable.

```{r mr_trash_wheel_df}
mr_trash_wheel_df =
  read_excel(
    "./Data/trash_wheel_data/202207 Trash Wheel Collection Data.xlsx",
    sheet = 1,
    range = "A2:M549",
    na = "") |>
  janitor::clean_names() |>
  rename(
    weight = weight_tons,
    volume = volume_cubic_yards) |>
  mutate(
    homes_powered = (weight*500)/30
  ) |>
  mutate(
    trash_wheel = "mr trash wheel"
  ) |>
  mutate(
    year = as.integer(year)
  )
```

Second, I will import and clean the Professor Trash Wheel dataset. Data cleaning for this dataset involves setting missing values, cleaning the variable names, renaming two existing variables, and creating two new variables.

```{r prof_trash_wheel_df}
prof_trash_wheel_df =
  read_excel(
    "./Data/trash_wheel_data/202207 Trash Wheel Collection Data.xlsx",
    sheet = 2,
    range = "A2:L96",
    na = "") |>
  janitor::clean_names() |>
  rename(
    weight = weight_tons,
    volume = volume_cubic_yards) |>
  mutate(
    homes_powered = (weight*500)/30
  ) |>
  mutate(
    trash_wheel = "professor trash wheel"
  )
```

Third, I will import and clean the Professor Trash Wheel dataset. Data cleaning for this dataset involves setting missing values, cleaning the variable names, renaming two existing variables, and creating two new variables.

```{r gwynnda_trash_wheel_df}
gwynnda_trash_wheel_df =
  read_excel(
    "./Data/trash_wheel_data/202207 Trash Wheel Collection Data.xlsx",
    sheet = 4,
    range = "A2:J108",
    na = "") |>
  janitor::clean_names() |>
  rename(
    weight = weight_tons,
    volume = volume_cubic_yards) |>
  mutate(
    homes_powered = (weight*500)/30
  ) |>
  mutate(
    trash_wheel = "gwynnda trash wheel"
  )
```

Finally, I will merge all three datasets by using `bind_rows` instead of another join or merge type because most of the column names are consistent across datasets and rows just need to be added to create one horizontally-longer dataset.

```{r merge problem 2}
merged_trash_wheel_df =
  bind_rows(mr_trash_wheel_df, prof_trash_wheel_df, gwynnda_trash_wheel_df) |>
  select(trash_wheel, everything())
```

Each dataset describes trash collected by different trash wheels around Baltimore, Maryland from May 2014 to July 2022. The first dataset in this problem, Mr. Trash Wheel, contains `r nrow(mr_trash_wheel_df)` rows or observations and `r ncol(mr_trash_wheel_df)` columns. The second dataset in this problem, Professor Trash Wheel, contains `r nrow(prof_trash_wheel_df)` rows or observations and `r ncol(prof_trash_wheel_df)` columns. The third dataset in this problem, Gwynnda Trash Wheel, contains `r nrow(gwynnda_trash_wheel_df)` rows or observations and `r ncol(gwynnda_trash_wheel_df)` columns. Each row represents one different dumpster that the trash wheel picks up trash from; thus, Mr. Trash Wheel collects trash from the most dumpsters, followed by Gwynnda Trash Wheel and Professor Trash Wheel. The columns for each dataset tell us the `date` of trash collection, `weight` (in tons) and `volume` (in cubic yards) of that dumpster's trash, and different types of trash that the trash wheels have collected, such as `plastic bottles` and `grocery bags`. The merged dataset with all three trash wheels contains `r nrow(merged_trash_wheel_df)` rows or observations and `r ncol(merged_trash_wheel_df)` columns. Altogether, the three trash wheels collected `r sum(pull(merged_trash_wheel_df, weight))` tons of trash and `r sum(pull(merged_trash_wheel_df, volume))` cubic pounds of trash. Professor Trash Wheel alone collected `r filter(merged_trash_wheel_df, trash_wheel == "professor trash wheel") |> pull(weight) |> sum()` tons of trash. Altogether, the three trash wheels collected `r filter(merged_trash_wheel_df, month == "July", year == 2021) |> pull(cigarette_butts) |> sum()` cigarette butts in July of 2021, while Gwynnda Trash Wheel alone collected `r filter(merged_trash_wheel_df, month == "July", year == 2021, trash_wheel == "gwynnda trash wheel") |> pull(cigarette_butts) |> sum()` cigarette butts in July of 2021.

# Problem 3

First, I will import and clean the dataset of baseline demographics. The goal of this dataset is to identify whether participants qualified for the study and were carriers.

```{r}
demographics_df =
  read.csv(
    "./Data/data_mci/MCI_baseline.csv",
    skip = 1,
    na = "."
  ) |>
  janitor::clean_names() |>
  mutate(
    sex =
      case_match(
        sex,
        0 ~ "male",
        1 ~ "female"),
        sex = as.factor(sex)
  ) |>
  mutate(
    apoe4 =
      case_match(
        apoe4,
        0 ~ "non carrier",
        1 ~ "carrier"),
        apoe4 = as.factor(apoe4)
  ) |>
    filter(is.na(age_at_onset))
```

Second, I will import and clean the dataset of longitudinally observed biomarker values.

```{r amyloid_df}
amyloid_df =
  read.csv(
    "./Data/data_mci/mci_amyloid.csv",
    skip = 1,
    na = c("NA", "Na")
  ) |>
  janitor::clean_names() |>
  rename(id = study_id)
```